Using functor categories to generate intermediate code with Agda

### 2025-05-13 draft.tex
draft completed

\newcommand{\bracket}[1]{\left\{ #1 \right\}}
\newcommand{\ang}[1]{\left\langle #1 \right\rangle}
\newcommand{\intp}[1]{\left\llbracket #1 \right\rrbracket}

@@

        \subsection{Category}
        \begin{definition}[Category]
            A \emph{category} $\mathcal{C}$ is specified by
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$ (e.g.\ $f : X \to Y$);
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Laws)}
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    For all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}
        \end{definition}

@@

            \begin{definition}[Opposite category]
                Given a category $\mathcal{C}$, its \emph{opposite category} $\mathcal{C}^{\textbf{op}}$ is specified by the following:
                \begin{itemize}
                    \item 
                        The objects of $\mathcal{C}^{\textbf{op}}$ are the same as those of $\mathcal{C}$;
                    \item 
                        For each $X, Y \in \textbf{obj}(\mathcal{C})$, the morphisms from $X$ to $Y$ in $\mathcal{C}^{\textbf{op}}$ are the morphisms from $Y$ to $X$ in $\mathcal{C}$;
                    \item 
                        The identity morphism on $X$ in $\mathcal{C}^{\textbf{op}}$ is the identity morphism on $X$ in $\mathcal{C}$;
                    \item 
                        The composition of morphisms in $\mathcal{C}^{\textbf{op}}$ is defined as the composition of morphisms in $\mathcal{C}$.
                \end{itemize}
            \end{definition}

@@

        \subsection{Functor}
        \begin{definition}[Functor]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by
            \begin{itemize}
                \item 
                    a function 
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have: $ F(\textbf{id}_X) = \textbf{id}_{F(X)}$
                \item
                    For all $X, Y, Z \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, we have: $F(g \circ f) = F(g) \circ F(f)$
            \end{itemize}
        \end{definition}

@@

        \subsection{Yoneda lemma}
        \begin{definition}[Yoneda functor]
            Given a category $\mathcal{C}$, the \emph{Yoneda functor} $\yo: \mathcal{C} \to \hat{\mathcal{C}}$ is defined as follows:
            \begin{itemize}
                \item 
                    For each $X \in \textbf{obj}(\mathcal{C})$, $\yo(X)$ is the functor $\mathcal{C}^{\textbf{op}} \to \mathcal{Set}$ defined as:
                    \begin{equation} \label{eq: yoneda_element}
                        \yo(X)(Y) = \mathcal{C}{(Y,X)}
                    \end{equation}
                    For all $Y \in \textbf{obj}(\mathcal{C})$;
                \item
                    For each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $\yo(f)$ is the morphism $\yo(X) \to \yo(Y)$ defined a natural transformation whose component at any given $Z \in \mathcal{C}^{\textbf{op}}$ is given by:
                    \begin{equation} \label{eq: yoneda_morphism}
                        \begin{aligned}
                            (\yo(f))_Z : \mathcal{C}{(Z,Y)} &\to \mathcal{C}{(Z,X)} \\
                            g &\mapsto g \circ f
                        \end{aligned}
                    \end{equation}
                    for all $Z \in \textbf{obj}(\mathcal{C})$.
                \end{itemize}

@@

\chapter{Implementation} \label{chap: implementation}
    \minitoc
    This chapter details the implementation of the compiler in Agda. Starting with a brief overview of the tools used, it then describes the directory structure and file dependencies. Then it discusses the specification of the source language, and the target language, which lead us to the problem of implementing a rigorous natural number subtraction and two different approaches. We will further see the difference between the two approaches and how they affect the implementation of the customised library and the implementation of the compiler.

@@

        Here the definition of the integers $\bZ$ is in the customised library. Please refer to~\secref{sec: lib} for more details.

@@

    \section{Target Language}
    The target language is an assembly-style intermediate language for a stack machine. It is defined with four stack-descriptor-indexed families of non-terminals: 
    \begin{itemize}
        \item 
            $\bracket{\textsf{L}_{sd}}$: left-hand sides
        \item 
            $\bracket{\textsf{S}_{sd}}$: simple right-hand sides
        \item
            $\bracket{\textsf{R}_{sd}}$: right-hand sides
        \item
            $\bracket{\textsf{I}_{sd}}$: instruction sequences
    \end{itemize}

@@

    It is defined in Agda as follows:
    \[\includegraphics[scale=1.2]{/target/stack_descriptor.pdf}\]
    Here definition of natural numbers is in the customised library. Please refer to~\secref{sec: lib} for more details.

@@

    \subsubsection{Order}
    The stack descriptor is ordered lexicographically with $\leq_s$ as follows:
    \[\ang{f, d} \leq_s \ang{f', d'} \Leftrightarrow f < f' \lor (f = f' \land d \leq d')\]

@@

    \subsubsection{Addition and subtraction of stack descriptors}
    We define addition and subtraction of stack descriptors as follows:
    \[\ang{f, d} \pm_s n = \ang{f, d \pm n} \text{ for } n \in \mathbb{N}\]
    Addition of stack displacement is defined in Agda as follows:
    \[\includegraphics[scale=1.2]{/target/stack_descriptor_add.pdf}\]
    For subtraction, we need to make sure that the subtraction is valid, i.e. the subtracted displacement is less than or equal to the displacement of the current stack descriptor. This leads us to the problem of defining the subtraction of natural numbers, which is explained in detail in~\secref{subsec: subtraction}

    \subsubsection{Properties of order and addition}
    We can use reflexivity and transitivity of the order $<$ and $\leq$ to prove that the order of stack descriptors $\leq_s$ is a preorder. The proof is as follows:
    \begin{proof}
        We need to show that $\leq_s$ is reflexive and transitive. 
        \begin{itemize}
            \item 
                Reflexivity: $sd \leq_s sd$ is trivially true since $f = f$ and $d \leq d$.
            \item
                Transitivity: if $\ang{f, d} \leq_s \ang{f', d'}$ and $\ang{f', d'} \leq_s \ang{f'', d''}$, we do the following case split:
                \begin{itemize}
                    \item 
                        If $f < f'$, $f' = f''$ and $d \leq d'$, then $f < f''$; 
                    \item 
                        If $f < f'$ and $f' < f''$, then $f < f''$ by transitivity of $<$;
                    \item
                        If $f = f'$, $d \leq d'$ and $f' < f''$, then $f < f''$;
                    \item
                        If $f = f'$, $d \leq d'$, $f' = f''$ and $d' \leq d''$, then $f = f''$ and $d \leq d''$ by transitivity of $\leq$. \footnote{Note that the properties of $=$ is not required in as in the Agda definition of $\leq_s$, we directly define $f'$ to be $f$ instead of defining a equivalence relation, thus equivalence can be checked directly via type checking.}
                \end{itemize}
        \end{itemize}
    \end{proof}

@@

    Proof above requires the proof of the properties of natural numbers, which is defined in~\secref{subsec: natural_numbers}.

    \subsection{Grammar}
    The grammar of the target language is defined as follows, given the current stack descriptor $sd = \ang{f, d}$, and new stack descriptor $sd'$ and $sd^v$:
    \[\begin{aligned}
        \bracket{\textsf{L}_{sd}} ::={}& sd^v \quad \text{when } sd^v \leq_s sd \\
                        &\mid \texttt{sbrs}  \\
        \bracket{\textsf{S}_{sd}} ::={}& \bracket{\textsf{L}_{sd}} \\
                        &\mid \texttt{lit} \bracket{\text{integer}} \\
        \bracket{\textsf{R}_{sd}} ::={}& \bracket{\textsf{S}_{sd}} \\
                        &\mid \bracket{ \text{unary operator} } \bracket{ \textsf{S}_{sd}}\\
                        &\mid \bracket{\textsf{S}_{sd}} \bracket{\textsf{binary operator}} \bracket{ \textsf{S}_{sd} } \\
        \bracket{\textsf{I}_{sd}} ::={}& \texttt{stop} \\
                        &\left.
                        \begin{aligned}
                        &|{\ } \bracket{\textsf{L}_{sd+\delta}} \gets \bracket{\textsf{R}_{sd}}[\delta] \text{ ; } \bracket{\textsf{I}_{sd+\delta}} \\
                        &|{\ } \texttt{if } \bracket{\textsf{S}_{sd}} \bracket{\text{relational operator}} \bracket{\textsf{S}_{sd}}[\delta] \\
                        & \hphantom{\mid {}} \texttt{ then } \bracket{\textsf{I}_{sd+\delta}} \texttt{ else } \bracket{\textsf{I}_{sd+\delta}} \\
                        &|{\ } \texttt{adjustdisp} [\delta] \text{ ; } \bracket{\textsf{I}_{sd+\delta}} \\
                        \end{aligned}
                        \quad \right\} \text{if } d + \delta \geq 0  \\
                        &{}\mid{} \texttt{popto } sd' \text{ ; } \bracket{\textsf{I}_{sd'}} \quad \text{when } sd' \leq_s sd \\
    \end{aligned}\]

@@

    \subsubsection{Instruction sequences}
    In the instruction sequences, $\delta$ as a displacement can be either positive or negative. To make a rigorous definition, we define $\delta$ as a natural number, and treat the positive and negative displacements as two different instructions, using addition and subtraction of stack descriptors respectively. Since the definition of negative displacement involves the rigorous definition of subtraction, we will discuss the subtraction of natural numbers and corresponding implementation in~\secref{subsec: subtraction}. The definition (except for the negative displacement) in Agda is as follows:
    \[\includegraphics[scale=1.2]{/target/instruction_inc.pdf}\]

@@

    \subsection{Definition from standard library}
    The following is a list of definitions (or their variants) from the standard library that are used in the customised library:
    \begin{itemize}
        \item 
            Natural numbers, integers and addition.
        \item
            Equality, congruence and substitution.
        \item
            Order of natural numbers, including $\leq$ and $<$.
        \item
            Some properties of order and addition.
    \end{itemize}

    The properties of order and addition contains the following terms summarised in the table below:
    \begin{table}[H]
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            $\mb{\mathsf{\mathord{\leq}\text{-}refl}}$ & $\forall n \in \bN, n \leq n$ \\
            \hline
            $\mb{\mathsf{\mathord{\leq}\text{-}trans}}$ & $\forall m, n, p \in \bN, m \leq n \land n \leq p \Rightarrow m \leq p$ \\
            \hline
            $\mb{\mathsf{\mathord{<}\text{-}trans}}$ & $\forall m, n, p \in \bN, m < n \land n < p \Rightarrow m < p$ \\
            \hline
            $\mb{\mathsf{\mathord{<}\mathord{\to}s\mathord{\leq}}}$ & $\forall m, n \in \bN, m < n \Rightarrow \textsf{suc } m \leq n$ \\
            \hline
            $\mb{\mathsf{< \to \leq}}$ & $\forall m, n \in \bN, m < n \Rightarrow m \leq n$ \\
            \hline
            $\mb{\mathsf{\mathord{\leq}\text{-}irrelevance}}$ & $\forall m, n \in \bN$, we consider all proofs of $m \leq n$ to be equal \\
            \hline
            $\mb{\mathsf{\mathord{+}\mathord{\to}\mathord{\leq}}}$ & $\forall m, n \in \bN, m \leq m + n$ \\
            \hline
            $\mb{\mathsf{\mathord{+}\mathord{\to}\mathord{\leq}^r}}$ & $\forall m, n \in \bN, m \leq n + m$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition}
        \label{tab: properties}
    \end{table}

    The detailed definitions are included in Appendix~[ref needed].

@@

    \subsubsection{Totality of order}
    It is useful to case split the order of natural numbers, i.e. for any $m, n \in \bN$, we have $m \leq n$ or $n \leq m$. This is defined in Agda by defining either case as an instance of the \mb{\textsf{Total}} type, and then showing that for any $m, n \in \bN$, we can construct a \mb{\textsf{Total}} type by case-split and \og{\textsf{with}}-construct. The proof is as follows:
    \[\includegraphics[scale=1.2]{/lib/order_totality.pdf}\]

@@

    \subsubsection{\textsf{Fin}-based approach}
    Instead of include the proof directly, we can define the type of the subtrahend as a type that depends on the minuend, where the type encodes the proof. There is a standard library in Agda on finite sets, where a type $\mb{\mathsf{Fin}}$ is defined as a type that depends on a natural number $n$, where $\mb{\mathsf{Fin}}\ n$ can be seen as the set of natural numbers less than $n$. The definition is as follows:
    \[\includegraphics[scale=1.2]{/lib/fin/fin.pdf}\]

    An intuitive explanation of the definition of $\mb{\mathsf{Fin}}$ is as follows:
    \begin{itemize}
        \item Base case: when $n = 0$, there is no element less than $0$, so we do not have any constructor that gives us an element of type $\mb{\mathsf{Fin}}\ 0$. 
        \item Inductive case: 
        \begin{figure}[H]
            \centering
            \includegraphics[width=0.35\textwidth]{fin_inductive.pdf}
            \caption{Inductive case of $\mb{\mathsf{Fin}}$}
            \label{fig: fin_inductive}
        \end{figure}
    \end{itemize}

@@

    \subsubsection{Comparison of the two approaches}
    The explicit approach carries proofs directly in the type system, which is more transparent but verbose. The $\textsf{Fin}$-based approach intially appears to be more elegant, as it encapsulates bounds check within a refined type. However, this elegance is superficial: constructing a term of type $\mb{\mathsf{Fin}}\ n$ still requires the same underlying proof of boundedness, shifting complexity to auxiliary conversions.

@@

    We have a functor $\intp{-} : \Theta \to \mathcal{K}$, that interprets the type constructor $\to$ as the exponential functor $\Rightarrow_s$.

    \subsection{Continuations}
    In Reynolds' model, we use $\mathsf{Compl}$ and $\mathsf{Intcompl}$ to represent command and integer continuations. Mathematically we have
    \[\mathsf{Compl } sd : \mathsf{SD} \to O \qquad \mathsf{Intcompl } sd : \bZ \to (\mathsf{SD} \to O)\]
    where $O$ is an unspecified domain of outputs.

>>>>

\newcommand{\bracket}[1]{\mathord{<} #1 \mathord{>}}
\newcommand{\ang}[1]{\left\langle #1 \right\rangle}
\newcommand{\intp}[1]{\left\llbracket #1 \right\rrbracket}

@@

        \subsection{Category}
        \begin{definition}[Category] \label{def: category}
            A \emph{category} $\mathcal{C}$ is specified by
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$ (e.g.\ $f : X \to Y$);
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Laws)}
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    For all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}
        \end{definition}

@@

            \begin{definition}[Opposite category] \label{def: opposite}
                Given a category $\mathcal{C}$, its \emph{opposite category} $\mathcal{C}^{\textbf{op}}$ is specified by the following:
                \begin{itemize}
                    \item 
                        The objects of $\mathcal{C}^{\textbf{op}}$ are the same as those of $\mathcal{C}$;
                    \item 
                        For each $X, Y \in \textbf{obj}(\mathcal{C})$, the morphisms from $X$ to $Y$ in $\mathcal{C}^{\textbf{op}}$ are the morphisms from $Y$ to $X$ in $\mathcal{C}$;
                    \item 
                        The identity morphism on $X$ in $\mathcal{C}^{\textbf{op}}$ is the identity morphism on $X$ in $\mathcal{C}$;
                    \item 
                        The composition of morphisms in $\mathcal{C}^{\textbf{op}}$ is defined as the composition of morphisms in $\mathcal{C}$.
                \end{itemize}
            \end{definition}

@@

        \subsection{Functor}
        \begin{definition}[Functor] \label{def: functor}
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by
            \begin{itemize}
                \item 
                    a function referred to as object mapping
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a function referred to as functorial mapping
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have: $ F(\textbf{id}_X) = \textbf{id}_{F(X)}$
                \item
                    For all $X, Y, Z \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, we have: $F(g \circ f) = F(g) \circ F(f)$
            \end{itemize}
        \end{definition}

@@

        \subsection{Yoneda lemma}
        \begin{definition}[Yoneda functor] \label{def: yoneda}
            Given a category $\mathcal{C}$, the \emph{Yoneda functor} $\yo: \mathcal{C} \to \hat{\mathcal{C}}$ is defined as follows:
            \begin{itemize}
                \item 
                    For each $X \in \textbf{obj}(\mathcal{C})$, $\yo(X)$ is the functor $\mathcal{C}^{\textbf{op}} \to \mathcal{Set}$ defined as:
                    \begin{equation} \label{eq: yoneda_element}
                        \yo(X)(Y) = \mathcal{C}{(Y,X)}
                    \end{equation}
                    For all $Y \in \textbf{obj}(\mathcal{C})$;
                \item
                    For each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $\yo(f)$ is the morphism $\yo(X) \to \yo(Y)$ defined a natural transformation whose component at any given $Z \in \mathcal{C}^{\textbf{op}}$ is given by:
                    \begin{equation} \label{eq: yoneda_morphism}
                        \begin{aligned}
                            (\yo(f))_Z : \mathcal{C}{(Z,Y)} &\to \mathcal{C}{(Z,X)} \\
                            g &\mapsto g \circ f
                        \end{aligned}
                    \end{equation}
                    for all $Z \in \textbf{obj}(\mathcal{C})$.
                \end{itemize}

@@

\chapter{Implementation} \label{chap: implementation}
    \minitoc
    This chapter details the implementation of the compiler in Agda. Starting with a brief overview of the tools used, it then describes the directory structure and file dependencies. Then it discusses the specification of the source language, and the target language, which lead us to the problem of implementing a rigorous natural number subtraction and two different approaches. It further reveals the difference between the two approaches and how they affect the implementation of the customised library. Finally a detailed implementation of the compiler is presented.

@@

@@

    \section{Target Language}
    The target language is an assembly-style intermediate language for a stack machine. It is defined with four stack-descriptor-indexed families of non-terminals: 
    \begin{itemize}
        \item 
            $\bracket{\textsf{L}_{sd}}$: left-hand sides
        \item 
            $\bracket{\textsf{S}_{sd}}$: simple right-hand sides
        \item
            $\bracket{\textsf{R}_{sd}}$: right-hand sides
        \item
            $\bracket{\textsf{I}_{sd}}$: instruction sequences
    \end{itemize}
    The intent of the indexing here is any instructions with the form $\bracket{\textsf{I}_{sd}}$ is an instruction sequence that can be executed when the current stack descriptor is $sd$.

@@

    It is defined in Agda as follows:
    \[\includegraphics[scale=1.2]{/target/stack_descriptor.pdf}\]

@@

    \subsubsection{Order}
    The stack descriptor is ordered lexicographically with $\leq_\mathsf{s}$ as follows:
    \[\ang{f, d} \leq_\mathsf{s} \ang{f', d'} \Leftrightarrow f < f' \lor (f = f' \land d \leq d')\]

@@

    \subsubsection{Addition and subtraction of stack descriptors}
    We define addition and subtraction of stack descriptors as follows:
    \[\ang{f, d} \pm_s n = \ang{f, d \pm n} \text{ for } n \in \mathbb{N}\]
    Addition of stack displacement is defined in Agda as follows:
    \[\includegraphics[scale=1.2]{/target/stack_descriptor_add.pdf}\]
    For subtraction, we need to make sure that the subtraction is valid, i.e. the subtracted displacement adjustment is less than or equal to the displacement of the current stack descriptor. This leads us to the problem of defining the subtraction of natural numbers, which is explained in detail in~\secref{subsec: subtraction}

    \subsubsection{Properties of order and addition}
    We can use reflexivity and transitivity of the order $<$ and $\leq$ to prove that the order of stack descriptors $\leq_\mathsf{s}$ is a preorder. The proof is as follows:
    \begin{proof}
        We need to show that $\leq_\mathsf{s}$ is reflexive and transitive. 
        \begin{itemize}
            \item 
                Reflexivity: $sd \leq_\mathsf{s} sd$ is trivially true since $f = f$ and $d \leq d$.
            \item
                Transitivity: if $\ang{f, d} \leq_\mathsf{s} \ang{f', d'}$ and $\ang{f', d'} \leq_\mathsf{s} \ang{f'', d''}$, we do the following case split:
                \begin{itemize}
                    \item 
                        If $f < f'$, $f' = f''$ and $d \leq d'$, then $f < f''$; 
                    \item 
                        If $f < f'$ and $f' < f''$, then $f < f''$ by transitivity of $<$;
                    \item
                        If $f = f'$, $d \leq d'$ and $f' < f''$, then $f < f''$;
                    \item
                        If $f = f'$, $d \leq d'$, $f' = f''$ and $d' \leq d''$, then $f = f''$ and $d \leq d''$ by transitivity of $\leq$. \footnote{Note that the properties of $=$ is not required in as in the Agda definition of $\leq_\mathsf{s}$, we directly define $f'$ to be $f$ instead of defining a equivalence relation, thus equivalence can be checked directly via type checking.}
                \end{itemize}
        \end{itemize}
    \end{proof}

@@

    Proofs here are based on the properties of natural numbers specified in the customised library. Please refer to~\secref{sec: lib} for more details.

    \subsection{Grammar}
    The grammar of the target language is defined as follows, given the current stack descriptor $sd = \ang{f, d}$, and new stack descriptor $sd'$ and $sd^v$:
    \begin{figure}[H]
        \centering
        \[\begin{aligned}
            \bracket{\textsf{L}_{sd}} ::={}& sd^v \quad \text{when } sd^v \leq_\mathsf{s} sd \\
                            &\mid \texttt{sbrs}  \\
            \bracket{\textsf{S}_{sd}} ::={}& \bracket{\textsf{L}_{sd}} \\
                            &\mid \texttt{lit } \bracket{\text{integer}} \\
            \bracket{\textsf{R}_{sd}} ::={}& \bracket{\textsf{S}_{sd}} \\
                            &\mid \bracket{ \text{unary operator} } \bracket{ \textsf{S}_{sd}}\\
                            &\mid \bracket{\textsf{S}_{sd}} \bracket{\textsf{binary operator}} \bracket{ \textsf{S}_{sd} } \\
            \bracket{\textsf{I}_{sd}} ::={}& \texttt{stop} \\
                            &\left.
                            \begin{aligned}
                            &|{\ } \bracket{\textsf{L}_{sd+\delta}} := \bracket{\textsf{R}_{sd}}[\delta] \text{ ; } \bracket{\textsf{I}_{sd+\delta}} \\
                            &|{\ } \texttt{if } \bracket{\textsf{S}_{sd}} \bracket{\text{relational operator}} \bracket{\textsf{S}_{sd}}[\delta] \\
                            & \hphantom{\mid {}} \texttt{ then } \bracket{\textsf{I}_{sd+\delta}} \texttt{ else } \bracket{\textsf{I}_{sd+\delta}} \\
                            &|{\ } \texttt{adjustdisp} [\delta] \text{ ; } \bracket{\textsf{I}_{sd+\delta}} \\
                            \end{aligned}
                            \quad \right\} \text{if } d + \delta \geq 0  \\
                            &{}\mid{} \texttt{popto } sd' \text{ ; } \bracket{\textsf{I}_{sd'}} \quad \text{when } sd' \leq_\mathsf{s} sd \\
        \end{aligned}\]
        \caption{Target language grammar}
        \label{fig: target_grammar}
    \end{figure}
    Some explanations of the grammar are as follows:
    \begin{itemize}
        \item 
            $\texttt{sbrs}$: a register used to communicate the result of function procedures implemented by closed subroutines.
        \item 
            $\texttt{lit}$: a literal integer.
        \item 
            $\texttt{stop}$: stop the execution of the program.
        \item 
            $\delta$: a displacement adjustments, which can be any integer that satisfies the condition mentioned above.
        \item 
            $\texttt{adjustdisp}$: adjust the displacement of specified stack descriptor.
        \item 
            $\texttt{popto}$: pop to the specified stack descriptor, which is used to reduce the number of frames.
    \end{itemize}
    Note that the design of right-hand sides ensures that the right operand of the assignment contains at most one operator, which means that any instruction that involves multiple operators must be divided into multiple instructions. This feature of the target language significantly complicates the compilation of integer expressions, which we will analysis in detail in~\secref{subsubsec: compiler_intexp}.

@@

    \subsubsection{Instruction sequences}
    In the instruction sequences, $\delta$ as a displacement adjustment can be positive, zero or negative. To make a rigorous definition, we define $\delta$ as a natural number, and treat the positive and negative displacements as two different instructions, using addition and subtraction of stack descriptors respectively. Since the definition of negative displacement involves the rigorous definition of subtraction, we will discuss the subtraction of natural numbers and corresponding implementation in~\secref{subsec: subtraction}. The definition (except for the negative displacement) in Agda is as follows:
    \[\includegraphics[scale=1.2]{/target/instruction_inc.pdf}\]

@@

    \subsection{Definition from standard library}
    The following is a list of definitions (or their variants) from the standard library and PLFA~\cite{plfa} that are used in the customised library:
    \begin{itemize}
        \item 
            Natural numbers ($\mb{\bN}$), integers ($\mb{\bZ}$) and addition ($\mb{+}$).
        \item
            Equality ($\mb{\equiv}$), congruence ($\mb{\mathsf{cong}}$), substitution ($\mb{\mathsf{sub}}$) and transitivity of equality ($\mb{\mathsf{trans}}$).
        \item
            Order of natural numbers, including $\mb{\leq}$ and $\mb{<}$.
        \item
            Some properties of order and addition.
    \end{itemize}

    The properties of order and addition contains the following terms summarised in the table below:
    \begin{table}[H]
        \centering
        \begin{tabular}{|l|l|}
            \hline
            \textbf{Agda term} & \textbf{Mathematical meaning} \\
            \hhline{|=|=|}
            $\mb{+\text{-}\mathsf{identity^r}}$ & $\forall n \in \bN.\ n + 0 = n$ \\
            \hline
            $\mb{+\text{-}\mathsf{suc^r}}$ & $\forall m, n \in \bN.\ m + \textsf{suc } n = \textsf{suc } (m + n)$ \\
            \hline
            $\mb{+\text{-}\mathsf{comm}}$ & $\forall m, n \in \bN.\ m + n \equiv n + m$ \\
            \hline
            $\mb{\mathsf{\mathord{\leq}\text{-}irrelevant}}$ & For all $m, n \in \bN$, we consider all proofs of $m \leq n$ to be equal. \\
            \hline
            $\mb{\mathsf{\mathord{\leq}\text{-}refl}}$ & $\forall n \in \bN.\ n \leq n$ \\
            \hline
            $\mb{\mathsf{\mathord{\leq}\text{-}trans}}$ & $\forall m, n, p \in \bN.\ m \leq n \land n \leq p \Rightarrow m \leq p$ \\
            \hline
            $\mb{\mathsf{n\mathord{\leq}suc\text{-}n}}$ & $\forall n \in \bN.\ n \leq \textsf{suc } n$ \\
            \hline
            $\mb{\mathsf{m\mathord{\equiv}n,p\mathord{\leq}n\mathord{\to}p\mathord{\leq}m}}$ & $\forall p, m, n \in \bN.\ m \equiv n \land p \leq n \Rightarrow p \leq m$ \\
            \hline
            $\mb{\mathsf{\mathord{+}\mathord{\to}\mathord{\leq}}}$ & $\forall m, n \in \bN.\ m \leq m + n$ \\
            \hline
            $\mb{\mathsf{\mathord{+}\mathord{\to}\mathord{\leq}^r}}$ & $\forall m, n \in \bN.\ m \leq n + m$ \\
            \hline
            $\mb{\mathsf{\mathord{<}\text{-}trans}}$ & $\forall m, n, p \in \bN.\ m < n \land n < p \Rightarrow m < p$ \\
            \hline
            $\mb{\mathsf{\mathord{<}\mathord{\to}s\mathord{\leq}}}$ & $\forall m, n \in \bN.\ m < n \Rightarrow \textsf{suc } m \leq n$ \\
            \hline
            $\mb{\mathsf{< \to \leq}}$ & $\forall m, n \in \bN.\ m < n \Rightarrow m \leq n$ \\
            \hline
        \end{tabular}
        \caption{Properties of order and addition}
        \label{tab: properties}
    \end{table}

    The detailed definitions are included in Appendix~[TBD].

@@

    \subsubsection{Totality of order} \label{subsubsec: order_totality}
    It is useful to case split the order of natural numbers, i.e. for any $m, n \in \bN$, we have $m \leq n$ or $n \leq m$. This is defined in Agda by defining either case as an instance of the \mb{\textsf{Total}} type, and then showing that for any $m, n \in \bN$, we can construct a \mb{\textsf{Total}} type by case-split and \og{\textsf{with}}-construct. The proof is as follows:
    \[\includegraphics[scale=1.2]{/lib/order_totality.pdf}\]

@@

    \subsubsection{\textsf{Fin}-based approach}
    Instead of include the proof directly, we can define the type of the subtrahend to be dependent on the minuend, where the dependence encodes the proof. There is a standard library in Agda on finite sets, where a type $\mb{\mathsf{Fin}}$ is defined as a type that depends on a natural number $n$, where $\mb{\mathsf{Fin}}\ n$ can be seen as the set of natural numbers less than $n$. The definition is as follows:
    \[\includegraphics[scale=1.2]{/lib/fin/fin.pdf}\]

    An intuitive explanation of the definition of $\mb{\mathsf{Fin}}$ is as follows:
    \begin{itemize}
        \item Base case: when $n = 0$, we do not have any constructor that gives us an element of type $\mb{\mathsf{Fin}}\ 0$, corresponding to the empty set. 
        \item Inductive case: 
        \begin{figure}[H]
            \centering
            \includegraphics[width=0.5\textwidth]{fin_inductive.pdf}
            \caption{Inductive case of $\mb{\mathsf{Fin}}$}
            \label{fig: fin_inductive}
        \end{figure}
    \end{itemize}

@@

    \subsubsection{Comparison of the two approaches}
    The explicit approach carries proofs directly in the type system, which is more transparent but verbose. The $\textsf{Fin}$-based approach initially appears to be more elegant, as it encapsulates bounds check within a refined type. However, this elegance is superficial: constructing a term of type $\mb{\mathsf{Fin}}\ n$ still requires the same underlying proof of boundedness, shifting complexity to auxiliary conversions.

@@

    The definition of exponentials captures the essence of stack-respecting computations in the compiler. 
    \begin{itemize}
        \item $sd \leq_\mathsf{s} sd'$ means the new stack descriptor $sd'$ is larger than the old stack descriptor $sd$, representing the dynamic expansion of the stack during execution.
